Text Is All You Need

### 2023-05-11 docs/hes_final/presentation.qmd
tweak spike slide headings
Update statistical model slide

## The 'Spike' conflict metric

@@

### Spike Metric Cont...

@@

- In this approach we prepared the dataset by averaging news article embeddings per country, per month. We also included one-hot encoded columns for countries, but no other country features were added. 
- Given the high dimensionality of the data (512-dimensional embedding vectors + 700 dummy variables for countries), this dataset suffered from the “curse of dimensionality”.

::: {#fig-xgboost layout-ncol=1}
```{mermaid}
%%| fig-width: 5
%%| fig-height: 5
flowchart TD
    GDELT[GDELT Embeddings] --> |average by country by month| AGG(Monthly news representation)
    Countries[Country labels] --> |one-hot encoding| AGG

@@

    ACLED[Monthly ACLED fatality counts] --> |compute our 'spike' variable| Spike(Conflict response binary variable)
    Spike --> XGD
    Spike --> KNN
```
Statistical model architecture
:::

:::{.notes}
- This was handled by performing Principal Component Analysis; we found that reducing the number of dimensions to represent 80% of variance (~169 components with our selected train-test cutoff) improved the overall performance.
- Another issue was a class imbalance: conflict spike was present in only 14% of the training samples. We used random over-sampling technique to upsample the positive examples.
- The model used in this approach is a stacked model that averages probabilities independently predicted by a K-Nearest Neighbors model and an XGBoost model. We used standard open-source versions of the underlying models. The intention behind averaging the predicted probabilities was to reduce the impact of individual model weaknesses. 
:::

>>>>

## Our 'Spike' conflict metric

@@

## 'Spike' conflict metric example

@@

::: {.column width="50%"}
::: {#fig-xgboost layout-ncol=1}
```{mermaid}
%%| fig-width: 5
%%| fig-height: 5
flowchart TD
    GDELT[GDELT Embeddings] --> |average by country by month| AGG(Monthly news representation)
    Countries[Country labels] --> |one-hot encoding| AGG

@@

    ACLED[Monthly ACLED fatality counts] --> |compute our 'spike' variable| Spike(Conflict response binary variable)
    Spike --> XGD
    Spike --> KNN
```
Statistical model architecture
:::
:::

::: {.column width="50%"}
- Average article embeddings per country per month
- One-hot encoded country labels
- Reduce dimensionality with PCA
- Train two models: K Nearest Neighbors and XGBoost
- Average their predictions
:::

:::{.notes}
- In this approach we prepared the dataset by averaging news article embeddings per country, per month. We also included one-hot encoded columns for countries, but no other country features were added. 
- Given the high dimensionality of the data (512-dimensional embedding vectors + 700 dummy variables for countries), this dataset suffered from the “curse of dimensionality”. 
- This was handled by performing Principal Component Analysis; we found that reducing the number of dimensions to represent 80% of variance (~169 components with our selected train-test cutoff) improved the overall performance.
- Another issue was a class imbalance: conflict spike was present in only 14% of the training samples. We used random over-sampling technique to upsample the positive examples.
- The model used in this approach is a stacked model that averages probabilities independently predicted by a K-Nearest Neighbors model and an XGBoost model. We used standard open-source versions of the underlying models. The intention behind averaging the predicted probabilities was to reduce the impact of individual model weaknesses. 
:::
